Binocular integration of chromatic and luminance signals

### 2024-04-02 manuscript.qmd
Minor edits to text

---
title: "Binocular integration of chromatic and luminance signals"
author:
  - Daniel H. Baker
  - Kirralise J. Hansford
  - Federico G. Segala
  - Annie Y. Morsi
  - Rowan J. Huxley
  - Joel T. Martin
  - Maya Rockman
  - Alex R. Wade
format:
  pdf:
    keep-tex: true
pdf-engine: xelatex
# execute:
#   echo: false
bibliography: references.bib
csl: journal-of-vision.csl
header-includes: 
  \usepackage{float} \floatplacement{figure}{H} 
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{A\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}}
---

@@

A further property of 'blob' regions is that they are less strongly orientation-tuned than other regions of V1 [@Horton1981; @Livingstone1984], and biased towards low spatial frequencies [@Edwards1995; @Tootell1988]. Our recent work has investigated binocular combination for flickering discs of luminance, which are DC-balanced across time [@Segala2023]. Steady-state EEG responses from early visual cortex and psychophysical contrast matching data were both consistent with weak interocular suppression when using this stimulus arrangement. This is very different from the 'ocularity invariance' that is well-established when using DC-balanced periodic stimuli such as sine-wave gratings, and implies strong interocular suppression [@Meese2006; @Baker2017; @Moradi2009]. There is additional psychophysical evidence from matching studies that binocular combination can be close to linear for luminance increments [@Levelt1965; @Anstis1998], particularly against a dark background [@Baker2012]. Both structural and functional data therefore imply that binocular combination may differ between spatial and temporal contrast.

@@

All experiments were completed by the first author (DHB) and two additional participants, who differed for each experiment. Participants had no known abnormalities of binocular vision. Written informed consent was obtained before data collection began, and all procedures were approved by the ethics committee of the Department of Psychology at the University of York (ID number 2202).

@@

Psychometric functions from each experiment were fit using \emph{psignifit} 4 to estimate threshold and slope parameters via a Bayesian numerical integration method [@Schutt2016]. A cumulative Gaussian was used as the underlying function, and we converted the slope estimates ($\sigma$ parameters from the fitted Gaussians) to equivalent Weibull $\beta$ values using the approximation $\beta = 10.3/\sigma$.

@@

\begin{equation}
Stage1_R = \frac{C_L^m}{S + C_R + \omega C_L},
\end{equation}

@@

where $C_L$ and $C_R$ are the contrasts displayed to the left and right eyes, and $m$, $S$, $\omega$, $p$, $q$ and $Z$ are free parameters in the model. A further free parameter, $k$, represents additive internal noise, and is used to convert the model outputs to either d-prime or threshold values (note that in the original model specification [@Meese2006] this parameter was called $\sigma$, but we use the $k$ symbol here to avoid confusion with the standard deviation of the cumulative Gaussian used when fitting the psychometric functions to derive thresholds). Thresholds are defined by iteratively adjusting the target contrast until the following equality is satisfied:

@@

We ran the simplex algorithm from `r ntotalsimplexfits` random starting vectors for each data set, and chose the solution for each data set that gave the smallest RMS error.

@@

![Dipper functions and psychometric slopes from Experiment 1, averaged across three participants. Panels (a-c) show threshold data, and panels (d-f) represent the slope of the psychometric function expressed in Weibull $\beta$ units. Error bars give ±1SE across participants. Note that contrast values are expressed as a percentage of the maximum displayable contrast (see Procedures for details). Curves in panels (a-c) show the best fitting models, optimized using a simplex algorithm (described in section 4.4), and RMSE values give the root mean square errors of the fits.](Figures/dipperssimplex.pdf){#fig-dipperfig}

@@

Following @Meese2006, we also inspected the slope of the psychometric function for each condition (see [@fig-dipperfig]d-f). At detection threshold, slopes were relatively steep for all stimuli, with $\beta$ values around 4. As pedestal contrasts increased, slopes linearized and reduced to around $\beta=1.3$ [@Foley1981; @Meese2006], and remained shallow at high pedestal contrasts. The exception to this was the dichoptic condition, where slopes became extremely steep at high dichoptic mask contrasts, consistent with previous observations [@Meese2006; @Baker2013]. This was clear for all three data sets, with slope values in the range $4 < \beta < 8$. However, we observe that slope estimates are more variable than threshold estimates (note the large error bars), particularly when using adaptive staircases, which deploy the majority of trials close to threshold. Our second experiment therefore investigated the slope of the psychometric function in more detail for dichoptic masking using the method of constant stimuli.

@@

In our final experiment, we again measured dipper functions, but this time for a temporally modulating luminance disc. This was motivated by our recent work [@Segala2023] that appeared to show increased binocular facilitation and reduced interocular suppression for flickering disc stimuli (relative to gratings), measured using EEG and a psychophysical matching paradigm. The pattern of dipper functions for a 4Hz flickering disc (see [@fig-discdata]a) was very similar to that observed for achromatic gratings (see [@fig-dipperfig]a), and the binocular summation ratio at threshold was also similar (a factor of `r round(10^(summationDisc/20),digits=2)` for discs, vs `r round(10^(summation[1]/20),digits=2)` for gratings). Threshold data for individual participants are shown in [@fig-individualdiscs]. We also found a similar pattern of psychometric slope values ([@fig-discdata]b) as we had for gratings, though we note that the dichoptic condition did not produce the 'super-steep' psychometric functions we had observed in Experiments 1 & 2. Nevertheless, the dichoptic slopes are somewhat above those of the other pedestal arrangements at high contrasts.

@@

For consistency with previous work, we initially performed least-squares fits of the two-stage model [@Meese2006] for each of the chromaticity experiments, and the flickering disc experiment (7 free parameters per fit). The best model fits are shown by the curves in @fig-dipperfig & @fig-discdata, and provide an excellent description of the data, with RMS errors between `r round(min(allrms),digits=2)` and `r round(max(allrms),digits=2)`dB. Best fitting parameters are shown in the 'Simplex fits' section of @tbl-parametertable. We note that in previous work, the weight of interocular suppression ($\omega$ in the model) is implicitly fixed at 1. Here we allowed it to vary, but it still received a value of exactly 1 in all of our grating conditions, and slightly below 1 (`r round(allp[4,6],digits=2)`) for our flickering discs. The exponent values (*p* and *q*) for the red/green gratings are rather different from those of the other conditions. In previous work [@Meese2006] the second stage gain control nonlinearity (Equation 4) does typically have quite substantial exponent values, which balance the relatively mild nonlinearity at the first stage, and produce a compressive transducer that results in contrast masking (the handle of the dipper). The high value of *p* = `r round(allp[2,1], digits=2)` is therefore likely to be compensating for the low value of *m* = `r round(allp[2,3], digits=2)` at stage one, but may well represent the combination of several successive stages of nonlinearity, and perhaps also other phenomena such as uncertainty [@Pelli1985].

We additionally implemented a hierarchical Bayesian version of the model in order to estimate full posterior parameter distributions. This model was fit simultaneously to the full trial-by-trial data from all participants who participated in a given experiment (i.e. the model was fitted separately to each of the four dipper data sets). [@fig-bayesianmodel]a-d summarises the model behaviour, which displays the same pattern of dipper functions as we found empirically. Mean posterior parameter values are given in the lower rows of @tbl-parametertable. These correspond quite closely to the parameters from the simplex fitting and the original @Meese2006 parameters (reproduced in the first row of the table for reference). The panels in the right and upper margins of [@fig-bayesianmodel] show posterior distributions for each model parameter. Note in particular that the posterior distribution for the weight of interocular suppression ($\omega$, top right panel) overlaps 1 for each experiment, consistent with the strong dichoptic masking observed in the threshold data. We note that the distribution for $\omega$ in the flickering disc condition is somewhat lower than for the other three data sets. However this difference is not meaningful according to criteria such as comparing the 95% credible intervals of the distribution to a value of $\omega=1$.

@@

Finally we fitted an extended model, that included interocular suppression between the different pathways, to the data of Experiment 2. [@fig-MCSmodel] shows the model curves (panels a-c), which correspond closely to the data in [@fig-MCSfig]. One striking discrepancy is that the model predicts a region of negative d-prime for the case where the target and dichoptic mask have the same chromaticity (see the curved regions below 50\% correct). This feature was not generally present in our data, though it can be observed for one participant in [@fig-individualMCS]b,c. Our previous work [@Baker2013] has found evidence for this phenomenon (termed 'swan functions'), but it generally requires very high mask contrasts to be measurable empirically, and there may also be some individual differences; both factors might explain its absence here.

[@fig-MCSmodel]d,e shows the model estimates for the weight of interocular suppression in each combination of target and mask chromaticity. Although the weights might appear to indicate that (excluding within-mechanism effects) suppression is strongest between blue/yellow targets and achromatic masks (a weight of 0.56), this is somewhat complicated by the differences in sensitivity and mask contrast between conditions. We selected mask contrasts that were approximately equal multiples ($16\times$) of their own detection thresholds, based on the data of Experiment 1 ([@fig-dipperfig]a-c). However the modelling expressed these contrasts as a percentage of the maximum contrast that it was possible to display on the monitor (see Procedures). The values in brackets in [@fig-MCSmodel]d indicate the overall suppressive term on the denominator, calculated by multiplying the mask contrast by the fitted weight. These values indicate that the strongest suppression is instead between red/green targets and blue/yellow masks, consistent with the threshold elevation observed empirically ([@fig-MCSfig]g). In principle it would be possible to repeat this modelling using either cone contrast values, or contrast expressed relative to detection threshold, which would require re-fitting the model.

@@

Estimates of the binocular summation ratio at threshold fell within the range $\sqrt{2}$ to 2 for all four data sets. Consistent with previous reports [@Simmons2005], summation was slightly higher for chromatic versus achromatic stimuli, but fell short of perfect linear summation (a ratio of 2). In our model, summation at threshold is determined by the exponent at the first gain control stage (the *m* parameter in Equations 1 & 2). The summation ratio can be approximated by $2^{1/m}$ [see Figure 9 of @Baker2012], such that an exponent of $m=1$ produces a ratio of 2, and an exponent of $m = 2$ produces a ratio of $\sqrt{2}$. Both the simplex and Bayesian model fits (see @tbl-parametertable) generated parameter estimates in the range $1 < m < 1.4$, consistent with the high levels of summation observed empirically.

@@

Previous work has demonstrated that the slope of the psychometric function in 2AFC tasks can distinguish different types of masking, although it is much less widely reported than threshold measures. In particular, pedestal masking linearizes the slope [@Foley1981; @Meese2006], and within-channel dichoptic masking produces very steep slopes [@Meese2006; @Baker2013]. We replicate both of these effects here, and show that they extend to isoluminant chromatic pathways (see [@fig-dipperfig]d-f & @fig-MCSfig). We additionally show that dichoptic masking between different pathways does not produce unusually steep slopes (see [@fig-MCSfig]). It therefore more closely resembles other types of masking between visual channels, such as cross-orientation masking [@Meese2009], surround masking [@Yu2002], and masking from broadband noise [@Baker2012b; @Lu2008], which also do not impact psychometric slopes.

@@

Recent evidence indicates that the physiological substrate of interocular suppression may be neurons in layer 4 of primary visual cortex [@Dougherty2019]. Most cells in this layer are monocularly exciteable, in that their responses increase only by stimulation of their preferred eye. However, simultaneous stimulation of the non-preferred eye can modulate the response, usually in an inhibitory fashion, exactly as proposed at stage 1 of the two-stage model (Equations 1 & 2). In terms of perception, one consequence of this early suppression is to achieve 'ocularity invariance', whereby the perceived contrast of a stimulus viewed by one eye is equivalent to that of the same stimulus viewed by both eyes [@Baker2007]. Similar processes of invariance have also been reported using fMRI [@Moradi2009] and steady-state EEG [@Baker2017].

>>>>

---
title: "Binocular integration of chromatic and luminance signals"
author:
  - Daniel H. Baker
  - Kirralise J. Hansford
  - Federico G. Segala
  - Annie Y. Morsi
  - Rowan J. Huxley
  - Joel T. Martin
  - Maya Rockman
  - Alex R. Wade
format:
  pdf:
    keep-tex: true
  # docx:
  #   toc: false
  #   number-sections: true
# prefer-html: true
pdf-engine: xelatex
# execute:
#   echo: false
bibliography: references.bib
csl: journal-of-vision.csl
header-includes: 
  \usepackage{float} \floatplacement{figure}{H} 
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{A\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}}
---

@@

A further property of 'blob' regions is that they are less strongly orientation-tuned than other regions of V1 [@Horton1981; @Livingstone1984], and biased towards low spatial frequencies [@Edwards1995; @Tootell1988]. Our recent work has investigated binocular combination for flickering discs of luminance, which are DC-balanced across time [@Segala2023]. Steady-state EEG responses from early visual cortex and psychophysical contrast matching data were both consistent with weak interocular suppression when using this stimulus arrangement. This is very different from the 'ocularity invariance' (where binocular and monocular stimuli appear equal) that is well-established when using DC-balanced periodic stimuli such as sine-wave gratings, and implies strong interocular suppression [@Meese2006; @Baker2017; @Moradi2009]. There is additional psychophysical evidence from matching studies that binocular combination can be close to linear for luminance increments [@Levelt1965; @Anstis1998], particularly against a dark background [@Baker2012]. Both structural and functional data therefore imply that binocular combination may differ between spatial and temporal contrast.

@@

All experiments were completed by the first author (DHB) and two additional participants, who differed for each experiment. Participants had no known abnormalities of binocular or colour vision. Written informed consent was obtained before data collection began, and all procedures were approved by the ethics committee of the Department of Psychology at the University of York (ID number 2202).

@@

Psychometric functions from each experiment were fit using \emph{psignifit} 4 to estimate threshold and slope parameters via a Bayesian numerical integration method [@Schutt2016]. A cumulative Gaussian was used as the underlying function, with contrast values expressed in decibel (dB) units (where $C_{dB} = 20log_{10}(C_\%)$). We converted the slope estimates ($\sigma$ parameters from the fitted Gaussians) to equivalent Weibull $\beta$ values using the approximation $\beta = 10.3/\sigma$. Threshold was defined as the target contrast corresponding to an accuracy of 75% correct.

@@

\begin{equation}
Stage1_R = \frac{C_R^m}{S + C_R + \omega C_L},
\end{equation}

@@

where $C_L$ and $C_R$ are the contrasts displayed to the left and right eyes, and $m$, $S$, $\omega$, $p$, $q$ and $Z$ are free parameters in the model. A further free parameter, $k$, represents additive internal noise, and is used to convert the model outputs to either d-prime or threshold values (note that in the original model specification [@Meese2006] this parameter was called $\sigma$, but we use the $k$ symbol here to avoid confusion with the standard deviation of the cumulative Gaussian used when fitting the psychometric functions to estimate thresholds). Thresholds are defined by iteratively adjusting the target contrast until the following equality is satisfied:

@@

We ran the simplex algorithm from `r ntotalsimplexfits` random starting vectors for each data set, and chose the solution for each data set that gave the smallest root mean squared error (RMSE) between the model and data.

@@

![Dipper functions and psychometric slopes from Experiment 1, averaged across three participants. Panels (a-c) show threshold data, and panels (d-f) represent the slope of the psychometric function expressed in Weibull $\beta$ units. Error bars give ±1SE across participants. Note that contrast values are expressed as a percentage of the maximum displayable contrast (see Procedures for details). Curves in panels (a-c) show the best fitting models, optimized using a simplex algorithm, and RMSE values give the root mean squared errors of the fits.](Figures/dipperssimplex.pdf){#fig-dipperfig}

@@

Following @Meese2006, we also inspected the slope of the psychometric function for each condition (see [@fig-dipperfig]d-f). At detection threshold, slopes were relatively steep for all stimuli, with values around $\beta=4$. As pedestal contrasts increased, slopes linearized and reduced to around $\beta=1.3$ [@Foley1981; @Meese2006], and remained shallow at high pedestal contrasts. The exception to this was the dichoptic condition, where slopes became extremely steep at high dichoptic mask contrasts, consistent with previous observations [@Meese2006; @Baker2013]. This was clear for all three data sets, with slope values in the range $4 < \beta < 8$. However, we observe that slope estimates are more variable than threshold estimates (note the large error bars), particularly when using adaptive staircases, which deploy the majority of trials close to threshold. Our second experiment therefore investigated the slope of the psychometric function in more detail for dichoptic masking using the method of constant stimuli, as well as exploring dichoptic interactions between chromatic channels.

@@

In our final experiment, we again measured dipper functions, but this time for a disc temporally modulating in luminance. This was motivated by our recent work [@Segala2023] that appeared to show increased binocular facilitation and reduced interocular suppression for flickering disc stimuli (relative to gratings), measured using EEG and a psychophysical matching paradigm. The pattern of dipper functions for a 4Hz flickering disc (see [@fig-discdata]a) was very similar to that observed for achromatic gratings (see [@fig-dipperfig]a), and the binocular summation ratio at threshold was also similar (a factor of `r round(10^(summationDisc/20),digits=2)` for discs, vs `r round(10^(summation[1]/20),digits=2)` for gratings). Threshold data for individual participants are shown in [@fig-individualdiscs]. We also found a similar pattern of psychometric slope values ([@fig-discdata]b) as we had for gratings, though we note that the dichoptic condition did not produce the 'super-steep' psychometric functions we had observed in Experiments 1 & 2. Nevertheless, the dichoptic slopes are somewhat above those of the other pedestal arrangements at high contrasts.

@@

For consistency with previous work, we initially performed least-squares fits of the two-stage model [@Meese2006] for each of the chromaticity experiments, and the flickering disc experiment (7 free parameters per fit). The best model fits are shown by the curves in @fig-dipperfig and @fig-discdata, and provide an excellent description of the data, with RMS errors between `r round(min(allrms),digits=2)` and `r round(max(allrms),digits=2)`dB. Best fitting parameters are shown in the 'Simplex fits' section of @tbl-parametertable. We note that in previous work, the weight of interocular suppression ($\omega$ in the model) is implicitly fixed at 1. Here we allowed it to vary, but it still received a value of exactly 1 in all of our grating conditions, and slightly below 1 (`r round(allp[4,6],digits=2)`) for our flickering discs. The exponent values (*p* and *q*) for the red/green gratings are rather different from those of the other conditions. In previous work [@Meese2006] the second stage gain control nonlinearity (Equation 4) does typically have quite substantial exponent values, which balance the relatively mild nonlinearity at the first stage, and produce a compressive transducer that results in contrast masking (the handle of the dipper). The high value of *p* = `r round(allp[2,1], digits=2)` is therefore likely to be compensating for the low value of *m* = `r round(allp[2,3], digits=2)` at stage one, but may well represent the combination of several successive stages of nonlinearity, and perhaps also other phenomena such as uncertainty [@Pelli1985].

We additionally implemented a hierarchical Bayesian version of the model in order to estimate full posterior parameter distributions. This model was fit simultaneously to the full trial-by-trial data from all participants who participated in a given experiment (i.e. the model was fitted separately to each of the four dipper data sets). [@fig-bayesianmodel]a-d summarises the model behaviour, which displays the same pattern of dipper functions as we found empirically. Mean posterior parameter values are given in the lower rows of @tbl-parametertable. These correspond quite closely to the parameters from the simplex fitting and the original @Meese2006 parameters (reproduced in the first row of the table for reference). The panels in the right and upper margins of [@fig-bayesianmodel] show posterior distributions for each model parameter. Note in particular that the posterior distribution for the weight of interocular suppression ($\omega$, top right panel) overlaps 1 for each experiment, consistent with the strong dichoptic masking observed in the threshold data. We note that the distribution for $\omega$ in the flickering disc condition is somewhat lower than for the other three data sets. However this difference is not meaningful according to widely accepted criteria such as comparing the 95% intervals of the distribution to a value of $\omega=1$.

@@

Finally we fitted an extended model, that included interocular suppression between the different pathways, to the data of Experiment 2. [@fig-MCSmodel] shows the model curves (panels a-c), which correspond closely to the data in [@fig-MCSfig]. One striking discrepancy is that the model predicts a region of negative d-prime for the case where the target and dichoptic mask have the same chromaticity (see the curved regions below 50\% correct). This happens in the model because low contrast targets suppress the mask more than they excite the detecting mechanism, producing a net decrease in response.  The feature (termed a 'swan function') was not generally present in our empirical data, though it can be observed for one participant in [@fig-individualMCS]b,c. Our previous work [@Baker2013] has found evidence for this phenomenon, but it generally requires very high mask contrasts to be measurable empirically, and there may also be some individual differences; both factors might explain its absence here.

[@fig-MCSmodel]d,e shows the model estimates for the weight of interocular suppression in each combination of target and mask chromaticity. Although the weights might appear to indicate that (excluding within-mechanism effects) suppression is strongest between blue/yellow targets and achromatic masks (a weight of 0.56), this is somewhat complicated by the differences in mask contrast between conditions. We selected mask contrasts that were approximately equal multiples ($16\times$) of their own detection thresholds, based on the data of Experiment 1 ([@fig-dipperfig]a-c). However the modelling expressed these contrasts as a percentage of the maximum contrast that it was possible to display on the monitor (see Procedures). The values in brackets in [@fig-MCSmodel]d indicate the overall suppressive term on the denominator of the model at stage 1, calculated by multiplying the mask contrast by the fitted weight. These values indicate that the strongest suppression is instead between red/green targets and blue/yellow masks, consistent with the threshold elevation observed empirically ([@fig-MCSfig]g). In principle it would be possible to repeat this modelling using either cone contrast values, or contrast expressed relative to detection threshold, which would require re-fitting the model.

@@

Estimates of the binocular summation ratio at threshold fell within the range $\sqrt{2}$ to 2 for all four stimuli tested here. Consistent with previous reports [@Simmons2005], summation was slightly higher for chromatic versus achromatic stimuli, but fell short of perfect linear summation (a ratio of 2). In our model, summation at threshold is determined by the exponent at the first gain control stage (the *m* parameter in Equations 1 & 2). The summation ratio can be approximated by $2^{1/m}$ [see Figure 9 of @Baker2012], such that an exponent of $m=1$ produces a ratio of 2, and an exponent of $m = 2$ produces a ratio of $\sqrt{2}$. Both the simplex and Bayesian model fits (see @tbl-parametertable) generated parameter estimates in the range $1 < m < 1.4$, consistent with the high levels of summation observed empirically.

@@

Previous work has demonstrated that the slope of the psychometric function in 2AFC tasks can distinguish different types of masking, although it is much less widely reported than threshold measures. In particular, pedestal masking linearizes the slope [@Foley1981; @Meese2006], and within-channel dichoptic masking produces very steep slopes [@Meese2006; @Baker2013]. We replicate both of these effects here, and show that they extend to the isoluminant chromatic pathways (see [@fig-dipperfig]d-f & @fig-MCSfig). We additionally show that dichoptic masking between different pathways does not produce unusually steep slopes (see [@fig-MCSfig]). It therefore more closely resembles other types of masking between visual channels, such as cross-orientation masking [@Meese2009], surround masking [@Yu2002], and masking from broadband noise [@Baker2012b; @Lu2008], which also do not impact psychometric slopes.

@@

Recent evidence indicates that the physiological substrate of interocular suppression may be neurons in layer 4 of primary visual cortex [@Dougherty2019]. Most cells in this layer are monocularly exciteable, in that their responses increase only by stimulation of their preferred eye. However, simultaneous stimulation of the non-preferred eye can modulate the response, usually in an inhibitory fashion, exactly as proposed at stage 1 of the two-stage model (Equations 1 & 2). In terms of perception, one consequence of this early suppression is to achieve 'ocularity invariance', whereby the perceived contrast of a stimulus viewed by one eye is equivalent to that of the same stimulus viewed by both eyes [@Baker2007]. Similar processes of response invariance have also been reported using fMRI [@Moradi2009] and steady-state EEG [@Baker2017].
